Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-06 paper/paper.qmd
Almost finished first draft

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

# Old data (REMOVE ONCE THE TEXT ANALYSIS BIT IS UPDATED)
survey_data <- read_csv("../data/analysis_data/clean_STA302_postcourse_survey_w24.csv", show_col_types = FALSE)

# Read in data
all_data <- read_excel(here::here("data/raw_data/merged_data.xlsx"), sheet = "Sheet1") |>
  janitor::clean_names()
```

@@

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due students have received feedback on all their previous work in the class (including their past papers) and there is an, optional, two-day period of peer review.

@@

all_data |>
  mutate(mins_complete = completion_time - start_time) |> 
  filter(mins_complete < 1000) |> # Remove one student who took 4,000 minutes to complete
  # summarize(mean = mean(mins_complete),
            # standard_dev = sd(mins_complete))
  ggplot(aes(x = mins_complete)) +
  geom_histogram(binwidth = 1) +
  labs(
    x = "Time to complete survey (minutes)",
    y = "Frequency"
  ) +
  theme_minimal()

all_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(
    x = "Self-reported GPA",
    y = "Frequency"
  ) +
  theme_minimal()
```

@@

all_data |> 
  mutate(
    what_year_are_you = case_when(
      what_year_are_you == "PEY" ~ "4th", # Change PEY to 4th year
      what_year_are_you == "5th" ~ "5th or over",
      what_year_are_you == "6th" ~ "5th or over",
      TRUE ~ what_year_are_you
    )
  ) |> 
  count(what_year_are_you) |>
  pivot_wider(names_from = what_year_are_you, 
              values_from = n, 
              values_fill = list(n = 0)) |> 
  kable()
```

@@

writing_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(
    x = "Response",
    y = "Number of repondents",
    fill = "Response:"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )

@@

coding_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(
    x = "Response",
    y = "Number of repondents",
    fill = "Response:"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )
```

@@

all_data |>
  rename(ai_schoolwork_appropriate = to_what_extent_do_you_think_using_generative_ai_tools_such_as_chat_gpt_by_open_ai_or_equivalents_is_ethical_and_appropriate_for_schoolwork) |> 
  mutate(
    ai_schoolwork_appropriate = case_when(
      ai_schoolwork_appropriate == "Depends of what the main goal of schoolwork is" ~ "It depends",
      ai_schoolwork_appropriate == "I believe student can only use it if they are allowed to" ~ "It depends",
      ai_schoolwork_appropriate == "a controllable amount" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate depending on the context" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate with rules for usage" ~ "It depends",
      ai_schoolwork_appropriate == "depends" ~ "It depends",
      ai_schoolwork_appropriate == "depends on how it is used" ~ "It depends",
      ai_schoolwork_appropriate == "half and half" ~ "It depends",
      ai_schoolwork_appropriate == "unsure" ~ "It depends",
      TRUE ~ ai_schoolwork_appropriate
      )) |> 
  count(ai_schoolwork_appropriate) |> 
  kable(col.names = c("Ethical and appropriate for school?", "Number"))
```

To understand the role of LLMs in learning, students more granularly identified their usage by checking off various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among students in the course. More than half of students also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing.

@@

# Usefulness
all_data |>
  select(
    weekly_quiz,
    weekly_mini_essay,
    papers_generating_ideas,
    papers_writing_code,
    papers_writing_content,
    papers_improving_content
  ) |>
  pivot_longer(cols = everything(),
               names_to = "task",
               values_to = "response") |>
  mutate(
    # Simplify response categories and rename tasks
    response_simplified = case_when(
      response %in% c("\"Not helpful\"", "\"I did not use generative AI for this component\"") ~ "Less Helpful",
      response %in% c("\"Somewhat helpful\"", "\"Very helpful\"") ~ "More Helpful",
      TRUE ~ "Unknown"  # Handle unexpected values
    ),
    task = case_when(
      task == "weekly_quiz" ~ "Weekly Quiz",
      task == "weekly_mini_essay" ~ "Weekly Mini Essay",
      task == "papers_generating_ideas" ~ "Generating Ideas",
      task == "papers_writing_code" ~ "Writing Code",
      task == "papers_writing_content" ~ "Writing Content",
      task == "papers_improving_content" ~ "Improving Content",
      TRUE ~ "Other"  # Handle any unexpected task names
    )
  ) |> 
  group_by(task, response_simplified) |>
  summarise(count = n(), .groups = 'drop') |> 
  ggplot(aes(x = response_simplified, y = count, fill = response_simplified)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ task, scales = "free_y") +
  labs(y = "Number of respondents",
       x = "",
       fill = "Response type: ") +
  theme_minimal() +
  coord_flip() + 
  scale_fill_brewer(palette = "Set1") +
  theme(
    axis.text.y = element_blank(),
    legend.position = "bottom",
  )

```

@@

## Open-ended survey responses

```{r setup-sentiment}
#| message: false
#| echo: false
#| warning: false

#get positive words from tidytext
neg_words <- get_sentiments() |>
  filter(sentiment == "negative")

#get negative words from tidytext
pos_words <- get_sentiments() |>
  filter(sentiment == "positive")
```

One question asked students to elaborate on whether they thought generative AI tools such as ChatGPT were ethical and appropriate for schoolwork. This was an open-response question. We summarized the responses to that question using text analysis and the `tidytext` package [@tidytext].

```{r setup-ai-appropriate}
#| message: false
#| echo: false
#| warning: false

# AI Appropriateness Comments
ai_comments <- survey_data |>
  select(id, ai_schoolwork_appropriate_comments) |>
  mutate(comment = ai_schoolwork_appropriate_comments) |>
  select(-c(ai_schoolwork_appropriate_comments))

# put comments into tidytext format
tidy_ai_comments <- ai_comments |>
  group_by(id) |>
  unnest_tokens(word, comment) |>
  mutate(word = str_to_lower(string = word))

# Load sentiment lexicon
sentiments <- get_sentiments()

# Prepare data for comparison cloud
comparison_data <- tidy_ai_comments |>
  inner_join(sentiments) |>
  count(word, sentiment, sort = TRUE) |>
  acast(word ~ sentiment, value.var = "n", fill = 0)

# Example sentiments data frame
sentiments <- get_sentiments()

# Prepare data
recs_comments <- survey_data |>
  select(id, recs_for_ai_sta302) |>
  mutate(comment = recs_for_ai_sta302) |>
  select(-c(recs_for_ai_sta302))

# Tokenize and clean individual words
tidy_recs_comments <- recs_comments |>
  ungroup() |>
  unnest_tokens(word, comment) |>
  anti_join(stop_words)

# Count word frequencies
word_counts <- tidy_recs_comments |>
  count(word, sort = TRUE)

# Bigram analysis
bigrams <- recs_comments |>
  ungroup() |>
  unnest_tokens(bigram, comment, token = "ngrams", n = 2) |>
  separate(bigram, into = c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word &
           !word2 %in% stop_words$word) |>
  count(word1, word2, sort = TRUE) |>
  unite(bigram, word1, word2, sep = " ")

# Prepare data for comparison cloud (including both words and bigrams)
comparison_data_words <- tidy_recs_comments |>
  inner_join(sentiments, by = "word") |>
  count(word, sentiment, sort = TRUE)

comparison_data_bigrams <- bigrams |>
  separate(bigram, into = c("word1", "word2"), sep = " ") |>
  inner_join(sentiments, by = c("word1" = "word")) |>
  inner_join(sentiments,
             by = c("word2" = "word"),
             suffix = c(".word1", ".word2")) |>
  unite(bigram, word1, word2, sep = " ") |>
  count(bigram, sentiment.word1, sort = TRUE) |>
  rename(sentiment = sentiment.word1)

# Combine both data sets
combined_data <- bind_rows(
  comparison_data_words |> rename(term = word),
  comparison_data_bigrams |> rename(term = bigram)) |>
  arrange(desc(n))
```

```{r tbl-recs-relpolarity}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Comparing Incidences of Positive and Negative Words in Student.
#| tbl-subcap: ["Top 10 Positive Words", "Top 10 Negative Words"]
#| layout-ncol: 2
#| fig-align: center

# Generate positive words table
tidy_recs_comments |>
  ungroup() |>
  select(-c("id")) |>
  inner_join(pos_words) |>
  count(word, sort = TRUE) |>
  head(10) |>
  kable(, format = "markdown")

# Generate negative words table
tidy_recs_comments |>
  ungroup() |>
  select(-c("id")) |>
  inner_join(neg_words) |>
  count(word, sort = TRUE) |>
  head(10) |>
  kable(, format = "markdown") 
```

@@

all_data |>
  ggplot(mapping = aes(x = mark, color =)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  facet_wrap(vars(llm_usage),
             ncol = 2,
             labeller = labeller(
               llm_usage = c(
                 "Extensive" = "Extensive LLM usage",
                 "Minimal" = "Minimal LLM usage",
                 "None" = "No LLM usage",
                 "Somewhat" = "Some LLM usage"
               )
             )) +
  theme_minimal()
```

@@

As always with regression, average estimates over the full dataset. It may be that the results are different for different tranches of respondents. For instance, looking at the 28 students who received an A+ for the final paper, 13 of them had extensive LLM usage.

@@

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

@@

We don't know whether it made them faster.

@@

@@

```{r fig-posteriorchecks}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Posterior predictive checking
#| fig-subcap: ["Base model","Including self-reported GPA"]
#| layout-ncol: 2

@@

```{r fig-basemodeldiagnostics}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Base model diagnostics

@@

```{r fig-gpamodeldiagnostics}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Model including self-reported GPA diagnostics

>>>>

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

# Read in data
all_data <-
  read_excel(here::here("data/merged_data.xlsx"), sheet = "Sheet1") |>
  janitor::clean_names()
```

@@

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due students have received feedback on all their previous work in the class (including their past papers) and there is an optional two-day period of peer review.

@@

all_data |>
  mutate(mins_complete = completion_time - start_time) |>
  filter(mins_complete < 1000) |> # Remove one student who took 4,000 minutes to complete
  # summarize(mean = mean(mins_complete),
  # standard_dev = sd(mins_complete))
  ggplot(aes(x = mins_complete)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Time to complete survey (minutes)", y = "Frequency") +
  theme_minimal()

all_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(x = "Self-reported GPA", y = "Frequency") +
  theme_minimal()
```

@@

all_data |>
  mutate(
    what_year_are_you = case_when(
      what_year_are_you == "PEY" ~ "4th",
      # Change PEY to 4th year
      what_year_are_you == "5th" ~ "5th or over",
      what_year_are_you == "6th" ~ "5th or over",
      TRUE ~ what_year_are_you
    )
  ) |>
  count(what_year_are_you) |>
  pivot_wider(
    names_from = what_year_are_you,
    values_from = n,
    values_fill = list(n = 0)
  ) |>
  kable()
```

@@

writing_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

@@

coding_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")
```

@@

all_data |>
  rename(ai_schoolwork_appropriate = to_what_extent_do_you_think_using_generative_ai_tools_such_as_chat_gpt_by_open_ai_or_equivalents_is_ethical_and_appropriate_for_schoolwork) |>
  mutate(
    ai_schoolwork_appropriate = case_when(
      ai_schoolwork_appropriate == "Depends of what the main goal of schoolwork is" ~ "It depends",
      ai_schoolwork_appropriate == "I believe student can only use it if they are allowed to" ~ "It depends",
      ai_schoolwork_appropriate == "a controllable amount" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate depending on the context" ~ "It depends",
      ai_schoolwork_appropriate == "appropriate with rules for usage" ~ "It depends",
      ai_schoolwork_appropriate == "depends" ~ "It depends",
      ai_schoolwork_appropriate == "depends on how it is used" ~ "It depends",
      ai_schoolwork_appropriate == "half and half" ~ "It depends",
      ai_schoolwork_appropriate == "unsure" ~ "It depends",
      TRUE ~ ai_schoolwork_appropriate
    )
  ) |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical and appropriate for school?", "Number"))
```

To understand the role of LLMs in learning, students identified their usage in a more granular way by checking off various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among students in the course. More than half of students also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing.

@@

# Usefulness
all_data |>
  select(
    weekly_quiz,
    weekly_mini_essay,
    papers_generating_ideas,
    papers_writing_code,
    papers_writing_content,
    papers_improving_content
  ) |>
  pivot_longer(cols = everything(),
               names_to = "task",
               values_to = "response") |>
  mutate(
    # Simplify response categories and rename tasks
    response_simplified = case_when(
      response %in% c(
        "\"Not helpful\"",
        "\"I did not use generative AI for this component\""
      ) ~ "Less Helpful",
      response %in% c("\"Somewhat helpful\"", "\"Very helpful\"") ~ "More Helpful",
      TRUE ~ "Unknown"  # Handle unexpected values
    ),
    task = case_when(
      task == "weekly_quiz" ~ "Weekly Quiz",
      task == "weekly_mini_essay" ~ "Weekly Mini Essay",
      task == "papers_generating_ideas" ~ "Generating Ideas",
      task == "papers_writing_code" ~ "Writing Code",
      task == "papers_writing_content" ~ "Writing Content",
      task == "papers_improving_content" ~ "Improving Content",
      TRUE ~ "Other"  # Handle any unexpected task names
    )
  ) |>
  group_by(task, response_simplified) |>
  summarise(count = n(), .groups = 'drop') |>
  ggplot(aes(x = response_simplified, y = count, fill = response_simplified)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ task, scales = "free_y") +
  labs(y = "Number of respondents", x = "", fill = "Response type: ") +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.y = element_blank(), legend.position = "bottom")
```

@@

Finally, one question asked students to elaborate on whether they thought generative AI tools such as ChatGPT were ethical and appropriate for schoolwork. This was an open-response question. To provide a sense of the responses, we used Anthropic's Claude 3.5 Sonnet model (as at 5 August 2024) to summarize the comments and it provided:

> Many students view AI as a helpful supplementary tool, comparing it to resources like Google or calculators. They believe it can aid in understanding concepts, debugging code, brainstorming ideas, and saving time on routine tasks. However, there's a consensus that AI should not be used to complete entire assignments or replace original thinking. Students emphasize the importance of using AI ethically, citing it when appropriate, and not relying on it exclusively. Some argue that learning to use AI effectively is a valuable skill for future careers. Concerns raised include the potential for plagiarism, the risk of hindering critical thinking skills, and the possibility of receiving incorrect information. Overall, most students support the responsible use of AI in education, with proper guidelines and transparency, while recognizing the need to maintain academic integrity and develop independent learning skills.

@@

all_data |>
  ggplot(mapping = aes(x = mark, color = )) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  facet_wrap(vars(llm_usage),
             ncol = 2,
             labeller = labeller(
               llm_usage = c(
                 "Extensive" = "Extensive LLM usage",
                 "Minimal" = "Minimal LLM usage",
                 "None" = "No LLM usage",
                 "Somewhat" = "Some LLM usage"
               )
             )) +
  theme_minimal()
```

@@

@@

There is some very mild evidence from this observational study that LLM usage may be associated with better scores. However, this should mostly be treated as a direction for future research. More broadly, our work has three major take-aways: 1) LLMs have a major role to play in enhancing the trustworthiness of data science. 2) Making general purpose chatbots more opinionated and knowledgeable about data science would considerably help. 3) Alternative interaction approaches, beyond chatbots, could be especially useful.

## Impact of LLMs on trustworthy data science

## General purpose compared with opinionated specific chatbots

## On the need for LLM-based infrastructure

@@

There are a number of substantial weaknesses of this study. The foundational one is that we used observational data, much of which was self-reported. There may be selection bias present in terms of who used LLMs, who reported their LLM usage truthfully, and even who remained in the class. A different design, specifically a randomized controlled trial (RCT) would deal with many of these issues, although likely at some cost.

Regression reports average estimates over the full dataset. However, many earlier studies found distributional effects, with low-performing individuals benefiting more than high-performing individuals. Again, a change in design toward an RCT could enable the exploration of this question. Stratification of our dataset would result in small sample size, but nonetheless our data does provide some limited suggestive evidence that this effect may be present in data science. For instance, looking at the 28 students who received an A+ for the final paper, 13 of them had extensive LLM usage, whereas looking at the same number of worst performing students finds that only six of them had extensive LLM usage.

Along these lines, we have considered LLM usage for code and writing as equivalent, but they should actually have different impacts depending on student backgrounds. Distinguishing between native and second-language English speakers, and then focusing on differential LLM usage, could add a great deal to the analysis.

Finally, we only considered one outcome measure, namely grade on the final paper. Each paper required a considerable amount of time for the student to produce. If an LLM was found to reduce the time taken to produce a paper, without any reduction in quality, then that would be a similarly useful outcome.

Despite these shortcomings, our work clearly identifies a need for further research examining how LLMs can be used to develop a more trustworthy data science.

@@

We use `bayesplot` [@bayesplot] and `loo` [@loo] help with posterior predictive checks and model diagnostics.

@@

```{r fig-posteriorchecks}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Posterior predictive checking
#| fig-subcap: ["Base model", "Including self-reported GPA"]
#| layout-ncol: 2

@@

```{r fig-basemodeldiagnostics}
#| fig-height: 7
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Base model diagnostics

@@

```{r fig-gpamodeldiagnostics}
#| fig-height: 7
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Model including self-reported GPA diagnostics
